Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes.

### 2020-04-05 Ben_Manuscripts/hdphmm/Outline.tex
First pass of comments.

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.
  \begin{itemize}
    \item Using some base probability distribution, $H$, a Dirichlet process 
    (DP) generates discrete distributions, $G_0$, over a countably infinite 
    number of probability measures:
    \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
    \end{equation}
    where the $\theta_k$ are values drawn from the base distribution and the
    weights $\beta_k$ come from a stick-breaking process parameterized by the  
    concentration parameter $\gamma$ (equivalently referred to as GEM($\gamma$)). 
    \item Also expressed $G_0 \sim DP(\gamma, H)$
    %MRS3: This next line is a bit vauge (expresses confidence, closely related)
    %BJC3: tried to make more clear
    \item The concentration parameter, $\gamma$, expresses one's confidence in the 
    base distribution $H$. 
    \item When $\gamma\to 0$, the first weight of $G_0$, $\beta_1$, approaches 
    unity and for $\gamma\to\infty$, the weights become uniform and $G_0$ closely
    resembles $H$.
    \item Each row, $G_j$, of the transition matrix is produced by drawing from a DP specified 
    using the $\beta$ vector as a discrete base distribution and a separate concentration
    parameter, $\alpha$.
    \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
    \end{equation}
    \item This hierarchical specification ensures that the transition probabilities in 
    each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
    \item Once the model has converged only a finite number of states will have significant
    sampling.
  \end{itemize}

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the IHMM framework, we estimated the most likely number and sequence of hidden
  states in each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  \begin{itemize}
    \item We created a python implementation of this process which we heavily adapted from
    the MATLAB code of Fox et al.~\cite{fox_sticky_2007} 
    %BJC3: I got the software from this page: https://homes.cs.washington.edu/~ebfox/software/
    % Maybe I should reach out to emily fox to see how she would prefer I cite the software?
    %MRS3: be a bit more specific about what the criteria were for convergence.
    %BJC3: See if that's better
    \item Parameter estimation is iterative. Therefore, we looked for convergence 
    of each entry in the $\mathbf{c}$ vector as well as the $A$ and $\Sigma$ matrices of 
    Equation~\ref{eqn:var}.
    \item Plots illustrating parameter convergence are available in the SI.
    \item We refer the interested reader to much more extensive descriptions of 
    %MRS3: be more specific about which parts from the above are in the references
    %BJC3: made more specific
    the inference and sampling procedures used to estimate the VAR(1) parameters
    and the state sequence. 
    ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}
  \end{itemize}

@@

  %BJC2: I think it would be useful to continue using this trajectory to illustrate 
  % the parameterization process. It might be bunch of plots stacked on each other.
  %BJC2: More labels might be useful if not cluttered
  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{prior_guesses.pdf}
  \caption{The parameters of the prior on $c$ (black line) should be chosen such
  that the mean levels of each state identified in the trajectory (blue line) lie within
  regions of the prior with reasonable probability. We chose the mean of the prior 
  as halfway between the maximum and minimum of each trajectory dimension. We chose 
  $\sigma$ of the prior by defining the maximum and minimum to be 2 standard deviations
  from the mean (dashed lines).}\label{fig:prior_guesses}
  \end{figure}

@@

  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) because $x$ and $y$ aren't 
  meaningful alone due to the system's radial symmetry.
  \begin{itemize}
    \item We converted the $x$ and $y$ center-of-mass coordinates to $r$.
    \item We fixed the state sequence from the 3D parmeterization and applied the
    inference component of the IHMM procedure to the cylindrical trajectories in
    order to estimate the VAR(1) parameters in terms of $r$ and $z$.
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
  	\item For each solute studied, we identified 200-325 independent states, each
   	with separate VAR(1) parameters.
   	\item Many of these states exhibit very similar dynamical behavior except their
   	mean levels are different, especially in the axial direction where solutes could
   	get trapped along a broad and continuous range of $z$ coordinates. 
  \end{itemize}

@@

  % BJC2: this part may expand a bit if I try different clustering approaches. But
  % this is the one I'm going with for now.
  We clustered based on the two diagonal entries of $A$ and the two diagonal entries 
  of $\Sigma$.
  \begin{itemize}
   \item One can choose alternative clustering features, such as the eigenvalues of
   the $A$ and $\Sigma$ matrices, but we chose their diagonals because they can be
   easily translated to solute behavior.
   \item We also chose to cluster the diagonals of each matrix independently before
   combining the clusters to assign final state labels. 
   \item If we divide the total states into $m$ clusters based on the diagonal entries
   of $A$ and, separately, $n$ clusters based on the diagonal entries of $\Sigma$,
   then there are a total of $mn$ possible cluster combinations.
   \item This allows slightly more flexibility in the total number of states compared
   to clustering on all four features at once, since there can be up to, but not 
   necessarily exactly, $mn$ states. 
  \end{itemize}

@@

  We remapped the state sequence based on the cluster assignments and
  generated a state transition probability matrix, $T$.
  \begin{itemize}
   \item The IHMM algorithm also produces an estimate of $T$, but since we 
   fixed the state sequence, we decided to explicitly calculate $T$ by 
   counting the number of transitions between states compiled across all 
   trajectories.
   %MRS3: seems like you should just be able to lump together the IHMM states to generate the IHMM estimate of T?
   %BJC3: That's what I do. Maybe I'm misunderstanding your suggestion. I added a little bit to the sentence
   % above to try and make more clear.
  \end{itemize}

@@

  \subsection{Estimating Flux and Selectivity}
  %BJC: can make this discussion somewhat short since it's already in previous paper. 
  % Should probably just hit all the main equations quickly.

@@

  %BJC3: I'm not sure if it's worth talking about all of the solutes in this section.
  % there is a lot to talk about just with methanol (which is all I've done for now -- wanted to
  % get feedback on analysis approach).
  %BJC3: I see this more as a demonstration of the model. I think this work has higher
  % impact potential so I don't want it to be 15 pages.
  %BJC3: Although, it might be a good idea to include Urea or Acetic acid as a second example
  % molecule because their behavior is pretty different since their carbonyl group tend to 
  % associate with sodium a lot.

  %BJC3: obviously can use a better thesis.
  We showcase this modeling approach by example. 
  \begin{itemize}
    \item In Figure~\ref{fig:mechanism_map}, we plot a representative 2D methanol 
    trajectory in cylindrical coordinates.
    \item The coordinates are colored according to the cluster with which methanol's
    dynamics are most consistent, which implies that this particular methanol 
    trajctory exhibits 6 distinct dynamical behaviors. 
    \item The solute timeseries are coupled with color coded bars which describe the
    solute's local environment and physical interactions that it undergoes at that
    time frame.
    \item From the top, they describe whether the solute is associated with sodium,
    whether the solute is hydrogen bonding with the monomer as well as what part of
    the monomer it is hydrogen bonding to, the local number density surrounding the
    solute and the radial location of the solute with respect to the nearest pore
    center (this information is redundant but can be a useful visual aid). %BJC3: still debating removing that bar
  \end{itemize}

>>>>

  %BJC: I'm not sure how much of this I should include. This is pretty much 
  % a reproduction of Fox et al's explanation: https://ieeexplore.ieee.org/abstract/document/5563110
  The infinite-state HMM overcomes this drawback by placing a hierarchical
  Dirichlet process (HDP) prior on the transition probabilities.
  \begin{itemize}
    \item Using some base probability distribution, $H$, a Dirichlet process 
    (DP) generates discrete distributions, $G_0$, over a countably infinite 
    number of probability measures:
    \begin{equation}
      G_0 = \sum_{k=1}^{\infty} \beta_k \delta_{\theta_k} ~~ \theta_k \sim H, \beta \sim GEM(\gamma)
    \end{equation}
    where the $\theta_k$ are values drawn from the base distribution and the
    weights $\beta_k$ come from a stick-breaking process parameterized by the  
    concentration parameter $\gamma$ (equivalently referred to as GEM($\gamma$)). 
    \item Also expressed $G_0 \sim DP(\gamma, H)$
    %MRS3: This next line is a bit vauge (expresses confidence, closely related)
    %BJC3: tried to make more clear
    %MRS4: physically mention what the base distribution is?
   \item The concentration parameter, $\gamma$, expresses one's confidence in the 
    base distribution $H$. 
    \item When $\gamma\to 0$, the first weight of $G_0$, $\beta_1$, approaches 
    unity and for $\gamma\to\infty$, the weights become uniform and $G_0$ closely
    resembles $H$.
    \item Each row, $G_j$, of the transition matrix is produced by drawing from a DP specified 
    using the $\beta$ vector as a discrete base distribution and a separate concentration
    parameter, $\alpha$.
    \begin{equation}
      G_j = \sum_{k=1}^{\infty} \pi_{jk} \delta_{\theta_k} ~~ \pi_j \sim DP(\alpha, \beta)
    \end{equation}
    \item This hierarchical specification ensures that the transition probabilities in 
    each row share the same support points \{$\theta_1$, ..., $\theta_k$\}.
    \item Once the model has converged only a finite number of states will have significant
    sampling.
  \end{itemize}

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the IHMM framework, we estimated the most likely number and sequence of hidden
  states in each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  \begin{itemize}
    \item We created a python implementation of this process which we heavily adapted from
    the MATLAB code of Fox et al.~\cite{fox_sticky_2007} 
    %BJC3: I got the software from this page: https://homes.cs.washington.edu/~ebfox/software/
    % Maybe I should reach out to emily fox to see how she would prefer I cite the software?
    %MRS4: yes, she how she wants to cite it.
    %MRS3: be a bit more specific about what the criteria were for convergence.
    %BJC3: See if that's better
    %MRS4: But what quantitative criteria did you use to determine convergence?
    \item Parameter estimation is iterative. Therefore, we looked for convergence 
    of each entry in the $\mathbf{c}$ vector as well as the $A$ and $\Sigma$ matrices of 
    Equation~\ref{eqn:var}.
    \item Plots illustrating parameter convergence are available in the SI.
    \item We refer the interested reader to much more extensive descriptions of 
    the inference and sampling procedures used to estimate the VAR(1) parameters
    and the state sequence. 
    ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}
  \end{itemize}

@@

  %BJC2: I think it would be useful to continue using this trajectory to illustrate 
  % the parameterization process. It might be bunch of plots stacked on each other.
  %BJC2: More labels might be useful if not cluttered
  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{prior_guesses.pdf}
  %MRS4: remind people what $c$ physically represents here. 
  \caption{The parameters of the prior on $c$ (black line) should be chosen such
  that the mean levels of each state identified in the trajectory (blue line) lie within
  regions of the prior with reasonable probability. We chose the mean of the prior 
  as halfway between the maximum and minimum of each trajectory dimension. We chose 
  $\sigma$ of the prior by defining the maximum and minimum to be 2 standard deviations
  from the mean (dashed lines).}\label{fig:prior_guesses}
  \end{figure}

@@

  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) because $x$ and $y$ aren't 
  meaningful alone due to the system's radial symmetry.
  \begin{itemize}
    %MRS4: Do you need to convert the covariance to cylintrical coordinates here before clustereing? 
    \item We converted the $x$ and $y$ center-of-mass coordinates to $r$.
    \item We fixed the state sequence from the 3D parmeterization and applied the
    inference component of the IHMM procedure to the cylindrical trajectories in
    order to estimate the VAR(1) parameters in terms of $r$ and $z$.
  \end{itemize}

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
  	\item For each solute studied, we identified 200-325 independent states, each
   	with separate VAR(1) parameters.
   	\item Many of these states exhibit very similar dynamical behavior except their
   	mean levels are different, especially in the axial direction where solutes could
   	get trapped along a broad and continuous range of $z$ coordinates. 
  %MRS4: Maybe better to phrase above in terms of the symmetry in z. Why does it matter the range is continuous?
  \end{itemize}

@@

  % BJC2: this part may expand a bit if I try different clustering approaches. But
  % this is the one I'm going with for now.
  We clustered based on the two diagonal entries of $A$ and the two diagonal entries 
  of $\Sigma$.
  \begin{itemize}
   \item One can choose alternative clustering features, such as the eigenvalues of
   the $A$ and $\Sigma$ matrices, but we chose their diagonals because they can be
   easily translated to solute behavior.
   %MRS4: suggest translating them here into solute behavior for the reader!
   \item We also chose to cluster the diagonals of each matrix independently before
   combining the clusters to assign final state labels. 
   \item If we divide the total states into $m$ clusters based on the diagonal entries
   of $A$ and, separately, $n$ clusters based on the diagonal entries of $\Sigma$,
   then there are a total of $mn$ possible cluster combinations.
   \item This allows slightly more flexibility in the total number of states compared
   to clustering on all four features at once, since there can be up to, but not 
   necessarily exactly, $mn$ states. 
   %MRS4: so no clustering based on transition probablity, correct?
  \end{itemize}

@@

  We remapped the state sequence based on the cluster assignments and
  generated a state transition probability matrix, $T$.
  \begin{itemize}
   \item The IHMM algorithm also produces an estimate of $T$, but since we 
   fixed the state sequence, we decided to explicitly calculate $T$ by 
   counting the number of transitions between states compiled across all 
   trajectories.
  \end{itemize}

@@

  \subsection{Estimating Flux and Selectivity}
  %BJC: can make this discussion somewhat short since it's already in previous paper. 
  %MRS4: yes, can be short
  % Should probably just hit all the main equations quickly.

@@

  %BJC3: I'm not sure if it's worth talking about all of the solutes in this section.
  % there is a lot to talk about just with methanol (which is all I've done for now -- wanted to
  % get feedback on analysis approach).
  %BJC3: I see this more as a demonstration of the model. I think this work has higher
  % impact potential so I don't want it to be 15 pages.
  %BJC3: Although, it might be a good idea to include Urea or Acetic acid as a second example
  % molecule because their behavior is pretty different since their carbonyl group tend to 
  %MRS4: there should be SOME summary information on the other molecules, but it can just be the conclusions.
  % associate with sodium a lot.

  %BJC3: obviously can use a better thesis.
  We showcase this modeling approach by example. 
  \begin{itemize}
    \item In Figure~\ref{fig:mechanism_map}, we plot a representative 2D methanol 
    trajectory in cylindrical coordinates.
    \item The coordinates are colored according to the cluster with which methanol's
    dynamics are most consistent, which implies that this particular methanol 
    trajctory exhibits 6 distinct dynamical behaviors. 
    \item The solute timeseries are coupled with color coded bars which describe the
    solute's local environment and physical interactions that it undergoes at that
    time frame.
    %MRS: maybe number (1-4) or label (a-d) the rows as well. 
    \item From the top, they describe whether the solute is associated with sodium,
    whether the solute is hydrogen bonding with the monomer as well as what part of
    the monomer it is hydrogen bonding to, the local number density surrounding the
    solute and the radial location of the solute with respect to the nearest pore
    center (this information is redundant but can be a useful visual aid). %BJC3: still debating removing that bar
  \end{itemize}
